Verifying a Postfix Language

### 2021-12-09 project/paper.tex
submitted

\title{Verifying a Postfix Language \\
    \large CS595 Final Project}
\author{Dustin Van Tate Testa and Andrew Neth}

\begin{document}
\maketitle
\begin{abstract}
A formalization and some proofs of soundness is given to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
\end{abstract}

\section{Introduction}
\subsection{Motivation}
WebAssembly is a typed, stack-machine based assembly language which is widely supported across all major browsers. A number of new languages have come about with the goal of providing better support for the platform. Additionally it has proven the utility of typed stack-machines as a reasonably safe way to write programs.

There appears to be little research into stack-based, functional languages (likely due to these languages not being common), thus more exploration on the topic could likely produce some value.

\href{https://github.com/dvtate/postfix-haskell}{The language this paper focuses on} is currently being used to power \href{https://ridder.xyz/game}{some insignificant demos}, but maybe in some distant future it will end up as the backend for a visual programming tool or some other application where guarantees of safety which could be produced here would be very important.

\subsection{The Language}
This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.

\subsubsection{Syntax}
\begin{itemize}
    \item \textbf{Primitive Literals:} The language supports numeric and string literals although in the paper we will likely only cover numbers.

@@

    In the language, identifiers can reference both typed and untyped values, but likely will be limited to typed values in the paper.

@@

    \texttt{(($..._1$)($..._2$ I32): Condition ) (: Action ) \$Identifier fun}.

    In the final paper this will likely get replaced with a ternary.

    \item \textbf{Other Operators:} Although the majority of primitive operators are defined in the standard library, the compiler actually has a few builtins as well. Although these are very important to actually using the language they aren't worthy of listing out in their entirety here. Some examples include: \texttt{require}, \texttt{use}, \texttt{import}, \texttt{export}, \texttt{asm}, \texttt{namespace}, \texttt{type}, \texttt{unpack}, \texttt{class}, \texttt{=}, \texttt{|}, \texttt{@}, \texttt{\~}, \texttt{==}, \texttt{make}
\end{itemize}

@@

\subsection{Included}

\subsection{Number Literals}
Number literals are denoted with subscript indicating the numeric type they correspond with.
This mirrors the actual language which uses the C-style syntax for number literal type inference.

\subsubsection{Branching}
As mentioned before a ternary operator would be a reasonable subset of branching via function overloading. The tern operator is defined as an operator which

\begin{verbatim}
# Tern takes an I32 and two values of the same type
# and gives one of those two values
$tern ~ :type # :type - (I32 $A $A) ($A) Arrow

# As it is a subset we can define it in terms of functions
((I32 Any Any): type swap type == ) (: ( $c $a $b ) = b ) $tern fun
((I32 Any Any): type swap type == && ) (: ( $c $a $b ) = a ) $tern fun
\end{verbatim}

\subsubsection{Identifiers}
Identifiers are included, however, it was found that the notation for 
dereferencing an identifier $\kweid{x}~\texttt{~}$ was confusing and
thus the behavior was changed and thus the default behavior was
changed such that unescaped identifiers are dereferenced instead of 
called and values must be explicitly called via the $\texttt{@}$
operator.

\subsubsection{Macros}
Macros use a specific instance of the type annotations, notably where the
input types are specified and the output types are inferred. In order to 
enable recursion, a syntactic fixed-point combinator was provided to replace
the $\texttt{rec}$ syntax.

\subsubsection{Identical to base language}
\begin{enumerate}
    \item Tuples and tuple related operators
    \item type operator
    \item $\texttt{Arrow}$ type
    \item some math and comparison operators
\end{enumerate}

@@

\begin{verbatim}
(()(I32 F32): ... ) $get_num =
get_num $n =
(: n type I32 == ) (: 
    # How do we know that it's an i32 here?
) $branch fun
(: n type F32 == ) (: 
    # How do we know that it's an f32 here?
) $branch fun
branch
\end{verbatim}

The above example program illustrates the problem with the language.

\section{Small Step Semantics For Values}
\subsection{Notation}
To annotate stack values we used the notation of two values being written next to each other, as they would be written in the language itself. The values at the bottom of the stack (furthest to the left) are often irrelevant, and thus can be omitted with this notation. The superscript * notation (ie - $e^*$) is used to annotate 0 or more of the preceding. This is explicitly defined in the syntax but implicitly expanded in the semantics.

\subsection{Value and Step Judgement}
The step judgement states that expressions are evaluated from left to right until everything is a value. Values are defined further as we go through.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Derivation of e* -> e*'
        \Rule{STEP}
            {e_1 \step e_1'
                \\v_i \val ~ \forall~ v_i \in v^*}
            {v^* ~ e_1 ~ ... ~ e_n \step v^* ~ e_1' ~ ... ~ e_n}
        \and

        % number literals
        \Rule{V-1}
             {\forall ~ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
             {\kwnumlit{n}{\tau} \val}
        \and

        % Tuple literals evaluation
        \Rule{TUPLES-1}
            {e^* \step e^{*\prime}}
            {( e^* ) \step ( e^{*\prime} )}
        \Rule{TUPLES-2}
            {v_i \val ~ \forall v_i \in v^*}
            {(v^*) \val}
        \and
    \end{mathpar}  
}

\subsection{Macros, Identifiers and Scoping}
The store, $\sigma$, consists of a list of \textit{scopes}, where each scope is a mapping from identifiers to values. Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope. A macro literal steps to a macro value by first stepping its input types until they're all values before substituting all of the values in the store within its body (forming a closure). Calling a macro value pushes a new empty mappings list onto the store, then places the macro's expressions and an $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}    
        % Closures capture lexically scoped variables via substitution 
        \Rule{LET}
            {v \val
                \\ \sigma = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n] ]
                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n, x_{n+1} \step v] ]}
            {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}

        \and \Rule{Identifier}
            { \kwid{x} \step v \in \sigma(-1)}
            {\sigma; \kwid{x} \step \sigma; v}

        \and \Rule{CALL-1}
            {\sigma' = \sigma \cup [~]
                \\ v = \kwmacro{\tau^*}{e^*}
                \\ v \val}
            {\sigma; v~ \textbf{@} \step \sigma'; e^* \kw{end\_scope}}

        \and \Rule{CALL-2}
            {\sigma = [s_1,...s_n,s_{n+1}] \\ \sigma' = [s_1,...s_n]}
            {\sigma; \kw{end\_scope} \step \sigma'; \varepsilon}

@@

\subsection{Others}
Notice the use of the \href{https://en.wikipedia.org/wiki/Kronecker_delta}{Kronecker Delta} to define the equals operator.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % unpack operator
        \Rule{TUPLES-3}
            {v_i \val ~\forall~ v_i \in v^*}
            {\kwunpack{(v^*)} \step v^*}

        % math
        \Rule{ADD}
            {v_1 \val \\ v_2 \val \\ v_3 = v_1 + v_2}
            {v_1~ v_2~ + \step v_3}
        \and
        \Rule{MUL}
            {v_1 \val \\ v_2 \val \\ v_3 = v_1 * v_2}
            {v_1~ v_2~ \times \step v_3}
        \and

        % Comparisons
        \Rule{EQUIV}
            {v_1 \val \\ v_2 \val}
            {v_1~ v_2~ ==~ \step \kwnumlit{(\delta_{v_1 v_2})}{\kwInt}}

        % Ternary
        \Rule{TERN-1}
            {v_1 \val \\ v_2 \val}
            {\kwtern{\kwnumlit{0}{\kwInt}}{v_1}{v_2} \step v_2}
        \Rule{TERN-2}
            {v_1 \val \\ v_2 \val \\ n \neq 0}
            {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}
        \and
    \end{mathpar}
}

@@

\subsection{Primitives}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{P-I32}{\strut}{\kwInt \prim} \and
        \Rule{P-I64}{\strut}{\kwLong \prim} \and
        \Rule{P-F32}{\strut}{\kwFloat \prim} \and
        \Rule{P-F64}{\strut}{\kwDouble \prim} \and
        %\Rule{P-sum}{\tau_1 \prim \\ \tau_2 \prim}{\tau_1 \tau_2 \texttt{|} \prim} \and
        \Rule{P-Arrow}{\tau_1 \prim \\ \tau_2 \prim}{\kwArrow{\tau_1}{\tau_2} \prim} \and
    \end{mathpar}  
}

\subsection{Step Judgement}
Type expressions are evaluated left to right until everything is $\prim$. Types and expressions have equal precedence and both can be present on the stack.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Derivation of T* -> T*'
        \Rule{TSTEP}
            {\tau \tstep \tau'
                \\\tau_i \prim ~ \forall~ \tau_i \in \tau_1^*}
            {\tau_1^* ~ \tau ~ \tau_2^* \tstep \tau_1^* ~ \tau' ~ \tau_2^*}

        \and \Rule{TSTEP - Mixed}
            {\tau^* \tstep \tau^{*\prime}
                \\ v_i \val ~ \forall v_i \in v^*}
            {v^*~ \tau^*~ e^*~ \tstep v^*~ \tau^{*\prime}~ e^*}
        \and \Rule{STEP - Mixed}
            {e^* \step e^{*\prime}
                \\ \tau_i \prim ~ \forall~ \tau_i \in \tau^*}
            {\tau^*~ e^* ~ \tau_2^*  \step \tau^* ~ e^{*\prime} ~ \tau_1}
    \end{mathpar}
}

\subsection{Type Identifiers}
Type identifiers use the same store as used for values, however their identifiers are guaranteed to not conflict as type identifiers start with a capital letter vs lower-case for value identifiers. Because we use the same store the step semantics for closures and scopes continue to work.
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{Type-Let}
            {\tau \prim
                \\ \sigma = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n] ]
                \\ \sigma' = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n, T_{n+1} \tstep \tau] ]}
            {\sigma; \tau ~ \kweid{T} = \tstep \sigma'; \varepsilon}

        \and \Rule{MACRO-2}
            {\sigma = [ ..., [x_1 \step v_1 ~ ... ~ x_n \step v_n, 
                ~ T_1 \tstep \tau_1 ~ ... ~ T_n \tstep \tau_n ]]
                \\ \tau_i \val~ \forall~ \tau_i \in \tau^*}
            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n][\tau_1/T_1]...[\tau_n/T_n] e^*} \val}

        \and \Rule{Type-Id}
            { \kwid{T} \step \tau \in \sigma(-1)}
            {\sigma; \kwid{T} \tstep \sigma; \tau}
    \end{mathpar}
}

\subsection{Tuples}
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        % Tuple type literals evaluation
        \Rule{Tuple Types - 1}
            {\tau^* \tstep \tau^{*\prime}}
            {( \tau^* ) \tstep ( \tau^{*\prime} )}
        \Rule{Tuple Types - 2}
            {\tau_i \prim ~ \forall \tau_i \in \tau^*}
            {(\tau^*) \prim}
        \and
        % unpack operator
        \Rule{Tuple Types - 3}
            {\tau_i \prim ~ \forall~ \tau_i \in \tau^*}
            {\kwunpack{(\tau^*)} \tstep \tau_1 ... \tau_n}
    \end{mathpar}
}

\subsection{Others}
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{Type Equiv}
            {\tau_1 \prim \\ \tau_2 \prim}
            {\tau_1~ \tau_2~ ==~ \tstep~ \kwnumlit{(\delta_{\tau_1 \tau_2})}{\kwInt}}

        % fixed point combinator
        \and \Rule{Fixed Point}
            {\kwmacro{\tau^*~ \tau_1}{e^*} \val \\ \tau_1 \prim}
            {\kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \val}
        \and \Rule{Call FP}
            {v = \kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \\ v \val}
            {v~ \textbf{@} \step v~ \kwmacro{\tau^*~ \tau_1}{e^*}~ \textbf{@}}

        % input types for macros can step
        \and \Rule{Macro Type Step}
            {\tau^* \tstep \tau^{*\prime}}
            {\sigma; \kwmacro{\tau^*}{e^*} \tstep \sigma; \kwmacro{\tau^{*\prime}}{e^*}}

        % typeof operator
        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \tstep \tau} 
    \end{mathpar}
}

@@

\subsection{Step Judgement}
{
    \centering
    \begin{mathpar}
    \Rule{T-0}
        {\typed{\ctx}{e}{\tau} \\ \tau \tstep \tau'}
        {\typed{\ctx}{e}{\tau'}}
    \end{mathpar}
}

\subsection{Math and Logic}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{T-1}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwInt}}{\kwInt}}
        \and \Rule{T-2}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwLong}}{\kwLong}}
        \and \Rule{T-3}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwFloat}}{\kwFloat}}
        \and \Rule{T-4}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwDouble}}{\kwDouble}}
        \and \Rule{T-5}
            {\typed{\ctx}{e_1}{\tau} 
                \\ \typed{\ctx}{e_2}{\tau} 
                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\typed{\ctx}{e_1~ e_2~ +}{\tau}}
        \and \Rule{T-6}
            {\typed{\ctx}{e_1}{\tau} 
                \\ \typed{\ctx}{e_2}{\tau} 
                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\typed{\ctx}{e_1~ e_2~ \times}{\tau}}
        \and \Rule{T-7}
            {\typed{\ctx}{e_1}{\kwInt}
                \\ \typed{\ctx}{e_2}{\tau}
                \\ \typed{\ctx}{e_3}{\tau}}
            {\typed{\ctx}{\kwtern{e_1}{e_2}{e_3}}{\tau}}
        \and \Rule{T-8}
            {\typed{\ctx}{e_1}{\tau}
                \\ \typed{\ctx}{e_2}{\tau}
                \\ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble \}}
            {\typed{\ctx}{e_1~ e_2~ ==~}{\kwInt}}
        \and \Rule{T-9}
            {\tau_1 \prim
                \\ \tau_2 \prim}
            {\typed{\ctx}{\tau_1~ \tau_2~ ==~}{\kwInt}}
    \end{mathpar}
}

\subsection{Tuples}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{Empty Tuple}{\strut}{\typed{\ctx}{()}{\kwUnit}}
        \and \Rule{Tuple Type}
            {\tau^* = \{ \tau_i ~\text{such that}~ \typed{\ctx}{e_i}{\tau_i}~ \forall ~ e_i \in e^* \} }
            {\typed{\ctx}{(e^*)}{(\tau^*)}}
        \and \Rule{Unpack}
            {\typed{\ctx}{(e^*)}{(\tau^*)}}
            {\typed{\ctx}{\kwunpack{(e^*)}}{\tau^*}}
    \end{mathpar}
}

\subsection{Macros}
This is why we needed the step judgement
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Move this to type rules   
        \Rule{Macro}
            {\tau^* = \tau_1 ~...~ \tau_n \\
             v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_n}{\tau_n} \\
             m = \kwmacro{\tau^*}{e^*} \\
             m \val }
            {\typed{\ctx}{\kwmacro{\tau^*}{e^*}}{\kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}}

        \and \Rule{Call}
            {\typed{\ctx}{e_1}{\kwArrow{\tau^*}{\tau^{*\prime}}}
                \\ \typed{\ctx}{e^*}{\tau^*}}
            {\typed{\ctx}{e^*~ e_1~ \textbf{@}}{\tau^{*\prime}}}

        \and \Rule{Identifier}
            {\ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{\kwid{x}}{\tau}}

>>>>

\title{Verifying a Postfix Language}
\author{Dustin Van Tate Testa and Andrew Neth}

\begin{document}
\maketitle
% \begin{abstract}
% A formalization and some proofs of soundness is given to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
% \end{abstract}

\section{Introduction}
\subsection{Motivation}
\href{https://github.com/dvtate/postfix-haskell}{The functional, postfix language this paper focuses on} is currently being used to power \href{https://ridder.xyz/game}{some insignificant demos}, but maybe in some distant future it will end up as the backend for a visual programming tool or some other application where guarantees of safety like the ones here would be very important.

\subsection{The Language}
This section describes some of the key features of the original language. It is included only to show that the subset we chose is a good representation of the original language and thus can be skipped.

\subsubsection{Full Syntax}
\begin{itemize}
    \item \textbf{Primitive Literals:} The language supports number and string literals

@@

@@

    \texttt{(($..._1$)($..._2$ I32): Condition ) (($..._1$): Action ) \$Identifier fun}.

    \item \textbf{Other Operators:} Although the majority of primitive operators are defined in the standard library, the compiler actually has a few builtins as well. Although these are very important to actually using the language, they aren't worthy of listing out in their entirety here. Some examples include: \texttt{require}, \texttt{use}, \texttt{import}, \texttt{export}, \texttt{asm}, \texttt{namespace}, \texttt{type}, \texttt{unpack}, \texttt{class}, \texttt{=}, \texttt{|}, \texttt{@}, \texttt{\~}, \texttt{==}, \texttt{make}
\end{itemize}

@@

\subsection{Included}
In addition to the following, tuples and tuple related operators,$\kwType{}$ operator, $\texttt{Arrow}$ type, and some math and comparison operators were included identically to their counterparts in base language.

\subsubsection{Number Literals}
Number literals are denoted with subscript indicating the numeric type they correspond with.
This mirrors the actual language which uses the C-style syntax for number literal type inference.

\subsubsection{Branching}
As mentioned before a ternary operator would be a reasonable subset of branching via function overloading. The $\kwtern{}{}{}$ operator is defined as an operator which takes a condition and two values of the same type, and gives one depending on the condition. Which could be defined like so:

\begin{verbatim}
((I32 Any Any): type swap type == ) (: ( $c $a $b ) = b ) $select fun
((I32 Any Any): type swap type == && ) (: ( $c $a $b ) = a ) $select fun
\end{verbatim}

\subsubsection{Identifiers}
Identifiers are included, however, it was found that the notation for 
dereferencing an identifier $\kweid{x}~\texttt{~}$ was confusing and
thus the subset was changed such that unescaped identifiers are dereferenced instead of 
called and values must be explicitly called via the $\texttt{@}$
operator.

\subsubsection{Macros}
Macros use a specific variation of the type annotations, notably where the
input types are specified and the output types are inferred. In order to 
enable recursion, a syntactic fixed-point combinator was provided to replace
the $\texttt{rec}$ syntax.

@@

\begin{verbatim}
(: type I32 == ) (: 
    # How do we know that it's an i32 here?
) $branch fun
(()(I32 F32): ... ) @ branch
\end{verbatim}

\section{Small Step Semantics For Values}
\subsection{Notation}
To annotate stack values we used the notation of two values being written next to each other, as they would be written in the language itself. The values at the bottom of the stack (furthest to the left) are often irrelevant, and thus for most rules can be omitted with this notation. The superscript * notation (ie - $e^*$) is used to annotate 0 or more of the preceding.

\subsection{Value and Step Judgement}
The step judgement states that expressions are evaluated from left to right until everything is a value. Values are defined further as we go through.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Derivation of e* -> e*'
        \Rule{Step}
            {e_1 \step e_1'
                \\v_i \val ~ \forall~ v_i \in v^*}
            {v^* ~ e_1 ~ ... ~ e_n \step v^* ~ e_1' ~ ... ~ e_n}
        \and

        % number literals
        \Rule{V-Num}
             {\forall ~ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
             {\kwnumlit{n}{\tau} \val}
        \and

        % Tuple literals evaluation
        \Rule{S-Tuple}
            {e^* \step e^{*\prime}}
            {( e^* ) \step ( e^{*\prime} )}
        \Rule{V-Tuple}
            {v_i \val ~ \forall v_i \in v^*}
            {(v^*) \val}
        \and
    \end{mathpar}  
}

\subsection{Macros, Identifiers and Scoping}
The store, $\sigma$, consists of a list of scopes, where each scope is a mapping from identifiers to values. Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope. A macro literal steps to a macro value by first stepping its input types until they're all values before substituting all of the values in the store within its body (forming a closure) (this rule comes later). Calling a macro value pushes a new empty mappings list onto the store, then places the macro's body and an $\kw{end\_scope}$ operator onto the stack, enabling locals.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}    
        % Closures capture lexically scoped variables via substitution 
        \Rule{S-Let}
            {v \val
                \\ \sigma = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n] ]
                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n, x_{n+1} \step v] ]}
            {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}

        \and \Rule{S-Ident}
            { \kwid{x} \step v \in \sigma(-1)}
            {\sigma; \kwid{x} \step \sigma; v}

        \and \Rule{S-Call}
            {\sigma' = \sigma \cup [~]
                \\ v = \kwmacro{\tau^*}{e^*}
                \\ v \val}
            {\sigma; v~ \textbf{@} \step \sigma'; e^* \kw{end\_scope}}

        \and \Rule{S-Fix}
            %{v = \kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \\ v \val}
            {f = \kwfix{v}{\tau} \\ f \val}
            %{v~ \textbf{@} \step v~ \kwmacro{\tau^*~ \tau_1}{e^*}~ \textbf{@}}
            {f~ \textbf{@} \step f~ v~ \textbf{@}}

        \and \Rule{S-End}
            {\sigma = [s_1,...s_n,s_{n+1}] \\ \sigma' = [s_1,...s_n]}
            {\sigma; \kw{end\_scope} \step \sigma'; \varepsilon}

@@

\subsection{Others}
Notice the use of the \href{https://en.wikipedia.org/wiki/Kronecker_delta}{Kronecker Delta} to define the equals operator. Here $\tau$ indicates the rule is defined for all numeric types $\tau$
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % unpack operator
        \Rule{S-Unpack}
            {v_i \val ~\forall~ v_i \in v^*}
            {\kwunpack{(v^*)} \step v^*}

        % math
        \Rule{S-Add}
            {\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ +~ \step \kwnumlit{(n_1 + n_2)}{\tau}}
        \and
        \Rule{S-Mul}
            {\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ \times~ \step \kwnumlit{(n_1 \times n_2)}{\tau}}
        \and

        % Comparisons
        \Rule{S-Eq}
            {\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ ==~ \step \kwnumlit{(\delta_{n_1 n_2})}{\kwInt}}

        % Ternary
        \Rule{S-Tern-0}
            {v_1 \val \\ v_2 \val}
            {\kwtern{\kwnumlit{0}{\kwInt}}{v_1}{v_2} \step v_2}
        \Rule{S-Tern-1}
            {v_1 \val \\ v_2 \val \\ n \neq \kwnumlit{0}{\kwInt}}            {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}
        \and
    \end{mathpar}
}

@@

\subsection{Primitives}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{P-Num}{\tau \in {\kwInt, \kwLong, \kwFloat, \kwDouble}}{\tau \prim} \and
        %\Rule{P-sum}{\tau_1 \prim \\ \tau_2 \prim}{\tau_1 \tau_2 \texttt{|} \prim} \and
        \Rule{P-Arrow}{\tau_1 \prim \\ \tau_2 \prim}{\kwArrow{\tau_1}{\tau_2} \prim} \and
    \end{mathpar}  
}

\subsection{Step Judgement}
Type expressions are evaluated left to right until everything is $\prim$. Types and expressions have equal precedence and both can be present on the stack.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Derivation of T* -> T*'
        \Rule{Ty-Step}
            {\tau \tstep \tau'
                \\\tau_i \prim ~ \forall~ \tau_i \in \tau_1^*}
            {\tau_1^* ~ \tau ~ \tau_2^* \tstep \tau_1^* ~ \tau' ~ \tau_2^*}

        \and \Rule{Ty-Step-Mixed}
            {\tau^* \tstep \tau^{*\prime}
                \\ v_i \val ~ \forall v_i \in v^*}
            {v^*~ \tau^*~ e^*~ \tstep v^*~ \tau^{*\prime}~ e^*}
        \and \Rule{Val-Step-Mixed}
            {e^* \step e^{*\prime}
                \\ \tau_i \prim ~ \forall~ \tau_i \in \tau^*}
            {\tau^*~ e^* ~ \tau_2^*  \step \tau^* ~ e^{*\prime} ~ \tau_1}
    \end{mathpar}
}

\subsection{Type Identifiers}
Type identifiers use the same store as used for values, however their identifiers are guaranteed to not conflict as type identifiers start with a capital letter vs lower-case for value identifiers. Because we use the same store the step semantics for closures and scopes continue to work.
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{Ty-Let}
            {\tau \prim
                \\ \sigma = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n] ]
                \\ \sigma' = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n, T_{n+1} \tstep \tau] ]}
            {\sigma; \tau ~ \kweid{T} = \tstep \sigma'; \varepsilon}

        \and \Rule{Macro-Subst}
            {\sigma = [ ..., [x_1 \step v_1 ~ ... ~ x_n \step v_n, 
                ~ T_1 \tstep \tau_1 ~ ... ~ T_n \tstep \tau_n ]]
                \\ \tau_i \val~ \forall~ \tau_i \in \tau^*}
            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n][\tau_1/T_1]...[\tau_n/T_n] e^*} \val}

        \and \Rule{Ty-Ident}
            { \kwid{T} \step \tau \in \sigma(-1)}
            {\sigma; \kwid{T} \tstep \sigma; \tau}
    \end{mathpar}
}

\subsection{Tuples}
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        % Tuple type literals evaluation
        \Rule{Ty-Tuple}
            {\tau^* \tstep \tau^{*\prime}}
            {( \tau^* ) \tstep ( \tau^{*\prime} )}
        \Rule{P-Tuple}
            {\tau_i \prim ~ \forall \tau_i \in \tau^*}
            {(\tau^*) \prim}
        \and
        % unpack operator
        \Rule{Ty-Unpack}
            {\tau_i \prim ~ \forall~ \tau_i \in \tau^*}
            {\kwunpack{(\tau^*)} \tstep \tau_1 ... \tau_n}
    \end{mathpar}
}

\subsection{Others}
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{S-TyEq}
            {\tau_1 \prim \\ \tau_2 \prim}
            {\tau_1~ \tau_2~ ==~ \step~ \kwnumlit{(\delta_{\tau_1 \tau_2})}{\kwInt}}

        % fixed point combinator
        \and \Rule{V-Fix}
            {e \val \\ \tau_1 \prim}
            {\kwfix{e}{\tau_1} \val}

        % input types for macros can step
        \and \Rule{Ty-Macro}
            {\tau^* \tstep \tau^{*\prime}}
            {\sigma; \kwmacro{\tau^*}{e^*} \tstep \sigma; \kwmacro{\tau^{*\prime}}{e^*}}

        % typeof operator
        \Rule{Ty-Typeof}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \tstep \tau} 
    \end{mathpar}
}

@@

\subsection{Step Judgement}
{
    \centering
    \begin{mathpar}
    \Rule{T-Step}
        {\typed{\ctx}{e}{\tau} \\ \tau \tstep \tau'}
        {\typed{\ctx}{e}{\tau'}}
    \end{mathpar}
}

\subsection{Math and Logic}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{T-Num}{\tau \in {\kwInt, \kwLong, \kwFloat, \kwDouble}}{\typed{\ctx}{\kwnumlit{n}{\tau}}{\tau}}
        \and \Rule{T-Add}
            {\typed{\ctx}{e_1}{\tau} 
                \\ \typed{\ctx}{e_2}{\tau} 
                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\typed{\ctx}{e_1~ e_2~ +}{\tau}}
        \and \Rule{T-Mul}
            {\typed{\ctx}{e_1}{\tau} 
                \\ \typed{\ctx}{e_2}{\tau} 
                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
            {\typed{\ctx}{e_1~ e_2~ \times}{\tau}}
        \and \Rule{T-Tern}
            {\typed{\ctx}{e_1}{\kwInt}
                \\ \typed{\ctx}{e_2}{\tau}
                \\ \typed{\ctx}{e_3}{\tau}}
            {\typed{\ctx}{\kwtern{e_1}{e_2}{e_3}}{\tau}}
        \and \Rule{T-Eq}
            {\typed{\ctx}{e_1}{\tau}
                \\ \typed{\ctx}{e_2}{\tau}
                \\ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble \}}
            {\typed{\ctx}{e_1~ e_2~ ==~}{\kwInt}}
        \and \Rule{T-TyEq}
            {\tau_1 \prim
                \\ \tau_2 \prim}
            {\typed{\ctx}{\tau_1~ \tau_2~ ==~}{\kwInt}}
    \end{mathpar}
}

\subsection{Tuples}
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{T-Unit}{\strut}{\typed{\ctx}{()}{\kwUnit}}
        \and \Rule{T-Tuple}
            {\tau^* = \{ \tau_i ~\text{such that}~ \typed{\ctx}{e_i}{\tau_i}~ \forall ~ e_i \in e^* \} }
            {\typed{\ctx}{(e^*)}{(\tau^*)}}
        \and \Rule{T-Unpack}
            {\typed{\ctx}{(e^*)}{(\tau^*)}}
            {\typed{\ctx}{\kwunpack{(e^*)}}{\tau^*}}
    \end{mathpar}
}

\subsection{Macros}
This is why we needed the step judgement
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Move this to type rules   
        \Rule{T-Macro}
            {\tau^* = \tau_1 ~...~ \tau_n \\
             v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_n}{\tau_n} \\
             m = \kwmacro{\tau^*}{e^*} \\
             m \val }
            {\typed{\ctx}{\kwmacro{\tau^*}{e^*}}{\kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}}

        \and \Rule{T-Call}
            {\typed{\ctx}{e_1}{\kwArrow{(\tau^*)}{(\tau^{*\prime})}}
                \\ \typed{\ctx}{e^*}{\tau^*}}
            {\typed{\ctx}{e^*~ e_1~ \textbf{@}}{\tau^{*\prime}}}

        \and \Rule{T-Ident}
            {\ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{\kwid{x}}{\tau}}

        \and \Rule{T-Let}
            {\typed{\ctx}{e}{\tau}
                \\ \ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{e~ \kwid{x} =}{\varepsilon}}

        \and \Rule{T-Fix}
            {\tau_1 \prim
                \\ \tau_1 = \kwArrow{(\tau^*_3)}{(\tau^*_4)}
                \\ \typed{\ctx}{e}{\kwArrow{(\tau^*_3 ~ \tau_1)}{(\tau^*_4)}}}
            {\typed{\ctx}{\kwfix{e}{\tau_1}}{\tau_1}}
